A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-05 tex/sec2-outline.tex
[ tex ] Revised abstract & S1

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take an untyped abstract syntax tree---which we call a \emph{raw term}---as input, and produce a typing derivation for the term if possible.
To give more precise definitions:
The raw terms%
\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
for simply typed $\lambda$-calculus are defined in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to help with type synthesis.
Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
%\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
Now we can define what it means to solve the type synthesis problem.

@@

Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.
Here we briefly recap the basic ideas of bidirectional type synthesis.
While the problem of type synthesis is not decidable in general, for certain kinds of terms it is still possible to synthesise their types --- for example, the $\SynRule{Var}$ rule says that the type of a variable can be looked up in the context.
For other kinds of terms, we can switch to the simpler problem of type \emph{checking}, where the expected type of a term is given as input so that there is more information to work with.
When a bidirectional type synthesiser traverses a raw term, it switches between the two \emph{modes}, synthesising wherever possible, and checking otherwise.
If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.

@@

One (smaller) benefit of mode preprocessing is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
Observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
That is, any term in synthesising mode is also in checking mode, but not necessarily vice versa.
A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
(We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so the term $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.

@@

The construction is straightforward:
Run the mode preprocessor on the input term~$|\Gamma| \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $|\Gamma| \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound} with respect to the original typing~(\cref{fig:STLC-typing-derivations}).

>>>>

We start with an overview of our theory by instantiating it to simply typed $\lambda$-calculus.
Roughly speaking, the problem of type synthesis requires us to take a raw term (i.e.~an untyped abstract syntax tree) as input, and produce a typing derivation for the term if possible.
To give more precise definitions:
The raw terms%
\footnote{We omit the usual conditions about named representations of variables throughout this paper.}
for simply typed $\lambda$-calculus are defined in \cref{fig:STLC-raw-terms};
besides the standard constructs, there is an \textsc{Anno} rule that allows the user to insert type annotations to help with type synthesis.
Correspondingly, the definition of typing derivations%
\footnote{We write~`$\bto$' instead of~`$\to$' for the function types of simply typed $\lambda$-calculus to avoid confusion with the function types in our type-theoretic meta-language.}
in \cref{fig:STLC-typing-derivations} also includes an \textsc{Anno} rule enforcing that the type of an annotated term does match the annotation.
%\footnote{The presence of type annotations allows type synthesis to subsume type checking: to check whether a term~$t$ has type~$A$, synthesise the type of $t \bbcolon A$.}
Now we can define what it means to solve the type synthesis problem.

@@

Now our goal is to use \cref{def:STLC-type-synthesiser} as a specification and implement it using a \emph{bidirectional} type synthesiser, which attempts to produce \emph{bidirectional} typing derivations defined in \cref{fig:STLC-bidirectional-typing-derivations}.

\Josh{Here we briefly recap the basic ideas of bidirectional type synthesis.
If a bidirectional type system is designed well (i.e.~\emph{mode-correct}, which will be defined in \cref{sec:type-synthesis}), then there will be a flow of type information in each rule that allows us to determine unknown types (e.g.~types to be checked) from known ones (e.g.~types previously synthesised), and suggests how bidirectional type synthesis is performed --- for example, the $\SynRule{App}$ rule says that to synthesise the type of an application $\isTerm{t\;u}$, we first synthesise the type of~$\isTerm{t}$, which should have the form $A \bto B$, from which we can extract the expected type of~$\isTerm{u}$, namely~$A$, and perform checking; then the type of the whole application, namely~$B$, can also be extracted from the previously synthesised type $A \bto B$.}

@@

One (less important) benefit of mode preprocessing is that it helps to simplify the synthesiser, whose computation can be partly directed by a mode derivation.
More importantly, whether there is a mode derivation for a term is actually very useful information to the user, because it corresponds to whether the term has enough type annotations:
Observe that the $\SynRule{Anno}$ and $\ChkRule{Sub}$ rules allow us to switch between the synthesising and checking modes;
the switch from synthesising to checking is free, whereas the opposite direction requires a type annotation.
That is, any term in synthesising mode is also in checking mode, but not necessarily vice versa.
A type annotation is required wherever a term that can only be in checking mode is required to be in synthesising mode, and a term does not have a mode derivation if and only if type annotations are missing in such places.
(We will treat all these more rigorously in \cref{sec:pre-synthesis}.)
For example, an abstraction is strictly in checking mode, but the left sub-term of an application has to be synthesising, so the term $\isTerm{(\lam{x}t)\;u}$ does not have a mode derivation unless we annotate the abstraction.

@@

The construction is straightforward:
Run the mode preprocessor on the input term~$|\Gamma| \vdash t$.
If there is no synthesising mode derivation, report that $t$~is not in synthesising mode (the third outcome).
Otherwise $|\Gamma| \vdash t^{\syn}$, and we can run the bidirectional type synthesiser.
If it finds a derivation of $\Gamma \vdash t \syn A$ for some type~$A$, return a derivation of $\Gamma \vdash t : A$ (the first outcome), which is possible because the bidirectional typing~(\cref{fig:STLC-bidirectional-typing-derivations}) is \emph{sound}%
\todo{Explain the notation `$:^\dir{d}$'}
with respect to the original typing~(\cref{fig:STLC-typing-derivations}).

### 2023-07-05 tex/sec4-pre-synthesis.tex
[ tex ] Compile Sec6 separately with lhs2TeX
[ tex ] Revised abstract & S1

\section{Soundness, Completeness, and Mode Preprocessing}\label{sec:pre-synthesis}

@@

@@

\subsection{Mode Preprocessing}
\label{sec:mode-preprocessing}

@@

\newcommand{\True}{\mathbf{T}}
\newcommand{\False}{\mathbf{F}}

@@

Intuitively, a term does not have a mode derivation exactly when there are not enough type annotations.
To formulate this more concretely, we can point out the places in the term that require annotations.
For a bidirectional type system, an annotation is required wherever a term is `strictly' (which we will explain shortly) in checking mode but required to be in synthesising mode, in which case there is no rule for switching from checking to synthesising, and thus there is no way to construct a mode derivation.
We can, however, consider \emph{generalised mode derivations} defined in \cref{fig:generalised-mode-derivations} that allow the use of an additional $\SynRule{Missing}$ rule for such switching, so that a derivation can always be constructed.
Given a generalised mode derivation, if it uses $\SynRule{Missing}$ in some places, then those places are exactly where annotations should be supplied, which is helpful information to the user; if it does not use $\SynRule{Missing}$, then the derivation is \emph{genuine} in the sense that it corresponds directly to an original mode derivation.
This can be succinctly formulated as \cref{thm:Pre?-true} below by encoding genuineness as a boolean~$g$ in the generalised mode judgement, which is set to~$\False$ only by the $\SynRule{Missing}$ rule.
(Ignore the boolean~$s$ for now.)

@@

\begin{theorem}\label{thm:generalised-mode-preprocessing}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
\end{theorem}

>>>>

\section{Mode Preprocessing and Related Properties}\label{sec:pre-synthesis}

@@

\Josh{One theorem to rule them all}

@@

\subsection{Generalised Mode Preprocessing}
\label{sec:mode-preprocessing}

@@

\renewcommand{\True}{\mathbf{T}}
\renewcommand{\False}{\mathbf{F}}

@@

Intuitively, a term does not have a mode derivation exactly when there are not enough type annotations, but such negative formulations convey little information.%
\todo{still default in the rest of the paper}
Instead, we can provide more information by pointing out the places in the term that require annotations.
For a bidirectional type system, an annotation is required wherever a term is `strictly' (which we will explain shortly) in checking mode but required to be in synthesising mode, in which case there is no rule for switching from checking to synthesising, and thus there is no way to construct a mode derivation.
We can, however, consider \emph{generalised mode derivations} defined in \cref{fig:generalised-mode-derivations} that allow the use of an additional $\SynRule{Missing}$ rule for such switching, so that a derivation can always be constructed.
Given a generalised mode derivation, if it uses $\SynRule{Missing}$ in some places, then those places are exactly where annotations should be supplied, which is helpful information to the user; if it does not use $\SynRule{Missing}$, then the derivation is \emph{genuine} in the sense that it corresponds directly to an original mode derivation.
This can be succinctly formulated as \cref{thm:Pre?-true} below by encoding genuineness as a boolean~$g$ in the generalised mode judgement, which is set to~$\False$ only by the $\SynRule{Missing}$ rule.
(Ignore the boolean~$s$ for now.)

@@

\begin{theorem}[Generalised Mode Preprocessing]\label{thm:generalised-mode-preprocessing}
For any raw term $V |-_{\Sigma,\erase\Omega} t$ and mode~$\dir{d}$, there is a derivation of\/ $V |-_{\Sigma,\Omega} t^{\dir{d}\,g\,s}$ for some booleans $g$~and~$s$.
\end{theorem}
